\documentclass[submit]{harvardml}

\course{CS181-S24}
\assignment{Assignment \#5}
\duedate{11:59pm EST, April 12, 2024}
\newcommand{\attr}[1]{\textsf{#1}}
\usepackage[OT1]{fontenc}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage[pdftex]{graphicx}
\usepackage{subfig}
\usepackage{framed}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{todonotes}
\usepackage{listings}
\usepackage{common}
\usepackage{bm}
\usepackage{tikz}
\usepackage{pythonhighlight}
\usetikzlibrary{positioning,shapes,arrows}

\usepackage[mmddyyyy,hhmmss]{datetime}

\definecolor{verbgray}{gray}{0.9}

\lstnewenvironment{csv}{%
  \lstset{backgroundcolor=\color{verbgray},
  frame=single,
  framerule=0pt,
  basicstyle=\ttfamily,
  columns=fullflexible}}{}

\begin{document}
\begin{center}
{\Large Homework 5: Mixtures, EM, and Graphical Models}\\
\end{center}

\subsection*{Introduction}

This homework assignment will have you work with EM for mixtures, PCA,
and graphical models.

\subsection*{Resources and Submission Instructions}

We encourage you to read sections 9.4 and 8.2.5 of the course textbook.

Please type your solutions after the corresponding problems using this
\LaTeX\ template, and start each problem on a new page.

Please submit the \textbf{writeup PDF to the Gradescope assignment `HW5'}. Remember to assign pages for each question.

Please submit your \textbf{\LaTeX\ file and code files to the Gradescope assignment `HW5 - Supplemental'}. 


\begin{problem}[Expectation-Maximization for Gamma Mixture Models: Derivations, 10pts]

In this problem we will explore expectation-maximization for a Categorical-Gamma Mixture model.

Let us suppose the following generative story for an observation $x$: first one of $K$ classes is randomly selected, and then the features $x$ are sampled according to this class. If $$z \sim \operatorname{Categorical}(\btheta)$$ indicates the selected class, then $x$ is sampled according to the class or ``component'' distribution corresponding to $z$. (Here, $\btheta$ is the mixing proportion over the $K$ components: $\sum_k \theta_k = 1$ and $ \theta_k > 0$). In this problem, we assume these component distributions are gamma distributions with shared shape parameter but different rate parameters: $$x | z \sim \operatorname{Gamma}(\alpha, \beta_k).$$

In an unsupervised setting, we are only given a set of observables as our training dataset: $\mathcal D = \{x_n\}_{n=1}^N$. The EM algorithm allows us to learn the underlying generative process (the parameters $\btheta$ and $\{\beta_k\}$) despite not having the latent variables $\{z_n\}$ corresponding to our training data.

\vspace{2em}

\begin{enumerate}

  \item \textbf{Intractability of the Data Likelihood} We are
    generally interested in finding a set of parameters $\beta_k$ that
    maximizes the likelihood of the observed data: $$\log
    p(\{x_n\}^N_{n=1}; \btheta, \{\beta_k\}^K_{k = 1}).$$ Expand the data
    likelihood to include the necessary sums over observations
    $x_n$ and to marginalize out the latents
    $\boldz_n$. Why is optimizing this likelihood directly
    intractable?

\item \textbf{Complete Data Log Likelihood} The complete dataset
  $\mathcal D = \{(x_n, \boldz_n)\}_{n=1}^N$ includes latents $\boldz_n$. Write
  out the negative complete data log likelihood: $$\mcL(\btheta, \{\beta_k\}^K_{k=1}) =  -\log p(\mathcal D; \btheta, \{\beta_k\}^K_{k=1}).$$

  Apply the power trick and simplify your expression using indicator elements $z_{n
  k}$.\footnote{The ``power trick'' is used when terms in a PDF are raised to the power of indicator components of a one-hot vector.  For example, it allows us to rewrite $p(\boldz_n ;  \btheta) = \prod_k \theta_k^{z_{nk}}$.} Notice that optimizing this loss is now computationally tractable if we know $\boldz_n$.

  (Continued on next page.)

\end{enumerate}

\end{problem}

\newpage


\begin{framed}
\noindent\textbf{Problem 1} (cont.)\\
\begin{enumerate}
\item[3.] \textbf{Expectation Step} Our next step is to introduce a
  mathematical expression for $\boldq_n$, the posterior over the
  hidden component variables~$\boldz_n$ conditioned on the observed data
  $x_n$ with fixed parameters.
That is:
  \begin{align*}
    \textbf{q}_n &= \begin{bmatrix}
      p(\boldz_n =\boldC_1| x_n; \btheta, \{ \beta_k \}^K_{k=1}) \\
      \vdots \\
      p(\boldz_n =\boldC_K| x_n; \btheta, \{ \beta_k \}^K_{k=1})
    \end{bmatrix}.
  \end{align*}
  %
%
  Write down and simplify the expression for
  $\boldq_n$.  Note that because the $\boldq_n$ represents the
  posterior over the hidden categorical variables $\boldz_n$, the
  components of vector $\boldq_n$ must sum to 1.
  The main work is to find an expression for $p(\boldz_n|x_n; \btheta, \{\beta_k\}^K_{k=1})$  for any choice of $\boldz_n$; i.e., for any 1-hot encoded $\boldz_n$. With this, you can then construct the different components that make up the vector $\boldq_n$.
  
\item[4.] \textbf{Maximization Step}
Using the~$\boldq_n$ estimates from the Expectation Step, derive an update for maximizing the expected complete data log likelihood in terms of $\btheta$ and $\{ \beta_k \}^K_{k=1}$.

\begin{enumerate}
    \item Derive an expression for the expected complete data log likelihood using $\boldq_n$.
    \item Find an expression for $\btheta$ that maximizes this expected complete data log likelihood. You may find it helpful to use Lagrange multipliers in order to enforce the constraint $\sum \theta_k = 1$. Why does this optimal $\btheta$ make intuitive sense?
    \item Find an expression for $\beta_k$ that maximizes the expected complete data log likelihood.  Why does this optimal $\beta_k$  make intuitive sense?
\end{enumerate}
    
\item[5.] Suppose that this had been a classification problem. That is,
  you were provided the ``true'' components $\boldz_n$ for each
  observation $x_n$,
  and you were going to perform the classification by
  inverting the provided generative model (i.e. now you're predicting $\boldz_n$ given $x_n$). Could you reuse any of
  your derivations above to estimate the parameters of the model?
  
\end{enumerate}
  
\end{framed}
\subsection*{Solution}

\begin{enumerate}
  \item 
  \item 
  \item 
  \item \begin{enumerate}
      \item 
      \item
      \item 
  \end{enumerate}
  \item 
\end{enumerate}


\newpage

\begin{problem}[Expectation-Maximization for Gamma Mixture Models: Coding, 15 pts]

  In this problem, you will implement your EM derivations from Problem
  1 and apply it to analyzing a synthetic example of the recovery time
  for patients following a surgical procedure, in hours.  The doctors
  have noticed that some patients seem to recover at an expected rate,
  but sometimes the recovery takes a long time.  They are keen to
  understand what is going on to improve their processes.  
  
  \begin{enumerate}

  \item Plot the data.  How would you describe the distribution?
    Based on what you see, why might a mixture model be an appropriate
    model?  
    
  \item Implement your solution from Problem 1 in \texttt{homework5.ipynb}.

{\bfseries You will recieve no points for code not included below.}

 
\item Run your code for 1, 2, 3, and 4 mixture components.  Plot the
  mixture models you find on top of the data distribution as well as
  the associated log likelihoods.  How many mixtures does it seem that
  there are?  How would you decide?
  
\item The doctors tell you that a normal recovery from the procedure
  is about 2-3 days, though sometimes patients recover a little
  faster.  Does this match what you see in the data?  Provide some
  hypotheses about what might be going on.

\item It's clear from the data that some patients take significantly
  longer than 2-3 days.  Do you observe that there is evidence that
  these represent a different cluster, vs. a long tail from a single
  cluster?  Why or why not?

\item The physician-scientists want to use this model to understand
  the characteristics of patients who have very long recoveries
  vs. those who do not.  Is this mixture modeling approach appropriate
  for this task?  Why or why not?

\item The physician-scientists develop a way of identifying someone's
  cluster based on a blood test---it seems that some patients in the
  longer group are ones that are at risk for clotting-related
  complications.  The hospital operations staff want to use this model
  to help streamline operations.  They plan to use the cluster of the
  patient to predict which patients will have a long length of stay.
  Is this plan sound?  May there be some issues?
  
  \end{enumerate} 
  
\end{problem} 

\subsection*{Solution}

\begin{enumerate}
  \item Plot:
  
  \item 
Code:

    \begin{python}
def e_step(theta, betas):
    q = 'not implemented'
    return q


def m_step(q):
    theta_hat = 'not implemented'
    beta_hats = 'not implemented'
    return theta_hat, beta_hats


def log_px(x, theta, betas):
    p = 'not implemented'
    return p


def run_em(theta, betas, iterations=1000):
    theta = 'not implemented'
    betas = 'not implemented'
    return theta, betas
    \end{python}
  \item Plots:
  \item 
  \item 
  \item 
  \item

\end{enumerate}


\newpage
  
\begin{problem}[PCA, 15 pts]

  
For this problem you will implement PCA from scratch on the first 6000 images of the MNIST dataset. Your job is to apply PCA on MNIST and discuss what kind of structure is found. Implement your solution in \texttt{homework5.ipynb} and attach the final plots below.

{\bfseries You will recieve no points for using third-party PCA implementations (i.e. {\normalfont \texttt{scikit-learn}}).}

{\bfseries You will receive no points for code not included below.}
\begin{enumerate}

\item Compute the PCA. Plot the eigenvalues corresponding to the most
  significant 500 components in order from most significant to
  least. Make another plot that describes the cumulative proportion of
  variance explained by the first $k$ most significant components for
  values of $k$ from 1 through 500.  How much variance is explained by
  the first 500 components?  Describe how the cumulative proportion of
  variance explained changes with $k$.  Include this plot below.

\item Plot the mean image of the dataset and plot an image
  corresponding to each of the first 10 principle components.  How do
  the principle component images compare to the cluster centers from
  K-means? Discuss any similarities and differences.  Include these
  two plots below.
  
  \textit{Reminder: Center the data before performing PCA}

\item Compute the reconstruction error on the data set using the mean
  image of the dataset.  Then compute the reconstruction error using
  the first 10 principal components.  How do these errors compare to
  the final objective loss achieved by using K-means on the dataset?
  Discuss any similarities and differences.

  For consistency in grading, define the reconstruction error as the squared L2
  norm averaged over all data points.


\item Suppose you took the original matrix of principle components
  that you found $U$ and multiplied it by some rotation matrix $R$.
  Would that change the quality of the reconstruction error in the
  last problem?  The interpretation of the components?  Why or why
  not?

\item Let's recall the zipcode application in Homework 3.  A common
  application of PCA is to dimensionality reduction before running a
  classifier: You first project the data onto the first few PCA bases,
  and then you train a classifier from the projection to the output.
  How might this be better than associating digits to clusters, as you
  did in Homework 4?  Would this approach help with any of the attacks
  in Homework 3?

\item You are collaborating with a penmanship analysis expert.  They
  are able to identify the kind of pen used to make a mark by various
  characteristics such as the width of the line, its crispness, and
  the type (if any) of ink splatter.  They have heard that your
  machine learning helped automate reading zip codes for the post
  office; they are wondering if you can help automate the manual
  process of classifying pen types.
  \begin{enumerate}
    \item Does what the expert is describing correspond to some kind
      of hidden representation or latent variable?  Describe why or
      why not.
    \item Do you think PCA will help the expert?  Why or why not?
      
  \end{enumerate}
\end{enumerate}
\end{problem}

\subsection*{Solution}
Code:

\begin{python}
def pca(x, n_comps=500):
    top_eigvals = 'not implemented'
    top_pcomps = 'not implemented'
    return top_eigvals, top_pcomps


def calc_cfvs(eigvals):
    cum_frac_vars = 'not implemented'
    return cum_frac_vars


def calc_errs(x, pcomps):
    err_mean = 'not implemented'
    err_pcomp = 'not implemented'
    return err_mean, err_pcomp
\end{python}

\begin{enumerate}
  \item Plots:
  \item Plots:
  \item 
  \item 
  \item 
  \item \begin{enumerate}
      \item 
      \item
  \end{enumerate}
\end{enumerate}

\begin{problem}[Bayesian Networks, 10 pts]

  \noindent In this problem we explore the conditional independence
  properties of a Bayesian Network.  Consider the following Bayesian
  network representing a fictitious person's activities. Each random
  variable is binary (true/false).

\begin{center}
  \begin{tikzpicture}[
    node distance=1cm and .5cm,
    bn/.style={draw,ellipse,text width=2cm,align=center}
      ]
      \node[bn] (w) {\attr{Weekend}};
      \node[bn,below right=of w] (t) {\attr{Traveling}};
      \node[bn,right=of t] (s) {\attr{Sick}};
      \node[bn,below left=of w] (f) {\attr{Friends over}};
      \node[bn,below right=of f] (eef) {\attr{Eat exotic foods}};
      \node[bn,right=of eef] (gs) {\attr{Get sleep}};
      \path (w) edge[-latex] (t)
      (w) edge[-latex] (f)
      (f) edge[-latex] (eef)
      (t) edge[-latex] (eef)
      (t) edge[-latex] (gs)
      (s) edge[-latex] (gs)
      (s) edge[-latex] (eef);
      \end{tikzpicture}
  \end{center}

  The random variables are:

  \begin{itemize}
  \item \attr{Weekend}: Is it the weekend?
  \item \attr{Friends over}: Does the person have friends over?
  \item \attr{Traveling}: Is the person traveling?
  \item \attr{Sick}: Is the person sick?
  \item \attr{Eat exotic foods}: Is the person eating exotic foods?
  \item \attr{Get Sleep}: Is the person getting sleep?
  \end{itemize}

  \medskip

  For the following questions, $A \perp B$ means that events A and B are
  independent and $A \perp B | C$ means that events A and B are independent
  conditioned on C.

  \textbf{Use the concept of d-separation} to answer the
  questions and show your work (i.e., state what the blocking path(s) is/are and what nodes block the path; or explain why each path is not blocked).

  \textit{Example Question:} Is $\attr{Friends over} \perp \attr{Traveling}$? If NO, give intuition for why.

  \textit{Example Answer:} NO. The path from Friends over -- Weekend -- Traveling is not blocked following the d-separation rules as we do not observe Weekend. Thus, the two are not independent. 

  \textbf{Actual Questions:}

\begin{enumerate}
\item Is $\attr{Weekend} \perp \attr{Get Sleep}$?
  If NO, give intuition for why.

\item Is $\attr{Sick} \perp \attr{Weekend}$?
  If NO, give intuition for why.


\item Is $\attr{Sick} \perp \attr{Friends over}\given \attr{Eat exotic
  foods}$? If NO, give intuition for why.


\item Is $\attr{Friends over} \perp \attr{Get Sleep}$? If NO, give
  intuition for why.

\item Is $\attr{Friends over} \perp \attr{Get Sleep} \given
  \attr{Traveling}$? If NO, give intuition for why.

\item Suppose the person stops traveling in ways that affect their
  sleep patterns.  Travel still
  affects whether they eat exotic foods.  Draw the modified network. (Feel free to reference the handout file for the commands for displaying the new network in \LaTeX).

\item For this modified network, is $\attr{Friends over} \perp
  \attr{Get Sleep}$? If NO, give an intuition why.  If YES,
  describe what observations (if any) would cause them to no longer be
  independent.

\end{enumerate}
\end{problem}

\subsection*{Solution}

\begin{enumerate}
  \item 
  \item 
  \item 
  \item 
  \item 
  \item
  \item
\end{enumerate}

\newpage


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Name and Calibration
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Name}

\subsection*{Collaborators and Resources}
Whom did you work with, and did you use any resources beyond cs181-textbook and your notes?

\end{document}
